---
title: 'Review of Data Science 1'
format:
  html: default
editor_options: 
  chunk_output_type: console
---


### Determinants of COVID vaccination rates

```{r}
#| include: false

library(tidyverse)
vaccine_data <- read_csv("https://joeroith.github.io/264_spring_2026/Data/vaccinations_2021.csv")
```


1. Consider only Minnesota and its surrounding states (Iowa, Wisconsin, North Dakota, and South Dakota).  We want to examine the relationship between the percentage who voted for Biden and the percentage of complete vaccinations by state.  Generate two plots to examine this relationship:

a) A scatterplot with points and smoothers colored by state.  Make sure the legend is ordered in a meaningful way, and include good labels on your axes and your legend.  Also leave off the error bars from your smoothers.

```{r}
vaccine_data |> 
  filter(state %in% c("Minnesota", "Iowa", "Wisconsin", "North Dakota", "South Dakota" ))|>
ggplot(aes(x=perc_Biden, y = perc_complete_vac, color = fct_reorder2(state, perc_Biden, perc_complete_vac))) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE)  +
  labs(x = "Percent votes for Biden", 
       y = "Percent of vacinations compltee",
       color = "State",
       title = "Graph")

```


b) One plot per state containing a scatterplot and a smoother.

```{r}
vaccine_data |>
   filter(state %in% c("Minnesota", "Iowa", "Wisconsin", "North Dakota", "South Dakota" ))|>
  ggplot(aes(x=perc_Biden, y = perc_complete_vac, color = state)) +
  geom_point() +
  geom_smooth(se = FALSE)  +
  facet_wrap(~state) +
  labs(x = "Votes for Biden", 
       y = "Percent of vacinations compltee",
       title = "Graph")
  
```


Describe which plot you prefer and why.  What can you learn from your preferred plot?
*Come back to this one*
I prefer plot b because is is less cluttered and esair to read and see the general tred of each state

4. Produce 3 different plots for illustrating the relationship between the rural_urban_code and percent vaccinated.  Hint: you can sometimes turn numeric variables into categorical variables for plotting purposes (e.g. `as.factor()`, `ifelse()`).

```{r}
vaccine_data |>
ggplot(aes(x = rural_urban_code, y = perc_complete_vac)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Rural–Urban Code",
    y = "Percent Vaccinated",
    title = "Vaccination Rates Urban vs Rural"
  )
```


```{r}

vaccine_data |>
ggplot(aes(x = as.factor(rural_urban_code), y = perc_complete_vac) ) +
  geom_boxplot() +
  labs(
    x = "Rural–Urban Code",
    y = "Percent Vaccinated",
    title = "Vaccination Rates Urban vs Rural")
```

```{r}
vaccine_data |>
mutate(urban_group = ifelse(rural_urban_code <= 5, "Urban", "Rural")) |>
ggplot(aes(x = urban_group, y = perc_complete_vac, fill = urban_group)) +
  geom_boxplot() +
  labs(
    x = "",
    y = "Percent Vaccinated",
    title = "Vaccination Rates Urban vs Rural")
```

My favorite graph is the last one where it is just grouped by urban and rural. It is better then the other 2 because in the first the data does not make much sense presented as a numeric data, and the second has too many boxes, in which it does not tell the reader if 1 is urban or rural.

alt text:

This is a box plot showing the trend in vaccination rates in urban vs rural county. There is 2 boxes for the rural and urban counties. On the y axis we have the percent of completed vaccination in that county ranging from 0 to 100. We can see that the mean vaccination percent for rural county's is around 40 while urban is around 50. This suggests that people on the more urban areas are more likely to be vaccinated.


State your favorite plot, why you like it better than the other two, and what you can learn from your favorite plot.  Create an alt text description of your favorite plot, using the Four Ingredient Model.  See [this link](https://drive.google.com/file/d/1BuEbL6vETBcFULmtlrwwZBUYhdEg0Zqb/view?usp=sharing) for reminders and references about alt text.


5. BEFORE running the code below, sketch the plot that will be produced by R.  AFTER running the code, describe what conclusion(s) can we draw from this plot?  

```{r}
#| eval: FALSE

vaccine_data |>
  filter(!is.na(perc_Biden)) |>
  mutate(big_states = fct_lump(state, n = 10)) |>
  group_by(big_states) |>
  summarize(IQR_Biden = IQR(perc_Biden)) |>
  mutate(big_states = fct_reorder(big_states, IQR_Biden)) |>
  ggplot() + 
    geom_point(aes(x = IQR_Biden, y = big_states))
```


6. In this question we will focus only on the 12 states in the Midwest (i.e. where region == "Midwest").

a) Create a tibble with the following information for each state.  Order states from least to greatest state population.

- number of different `rural_urban_code`s represented among the state's counties (there are 9 possible)
- total state population
- proportion of Metro counties
- median unemployment rate 

```{r}
(midwest <- vaccine_data |>
  filter(region == "Midwest") |>
  group_by(state) |>
  mutate(
    num_rural_codes = n_distinct(rural_urban_code),
    state_population = sum(tot_pop, na.rm = TRUE),
    prop_metro = mean(metro_status == "Metro", na.rm = TRUE),
    median_unemployment = median(unemployment_rate_2020, na.rm = TRUE)
  ) |>
  arrange(state_population))
```


b) Use your tibble in (a) to produce a plot of the relationship between proportion of Metro counties and median unemployment rate.  Points should be colored by the number of different `rural_urban_code`s in a state, but a single linear trend should be fit to all points.  What can you conclude from the plot?

```{r}
midwest |> 
  ggplot(aes(prop_metro, median_unemployment, color = num_rural_codes)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(x = "Proportion of metro counties",
       y = "Median Unemployment")
```



7. Generate an appropriate plot to compare vaccination rates between two subregions of the US: New England (which contains the states Maine, Vermont, New Hampshire, Massachusetts, Connecticut, Rhode Island) and the Upper Midwest (which, according to the USGS, contains the states Minnesota, Wisconsin, Michigan, Illinois, Indiana, and Iowa).  What can you conclude from your plot?


In this next section, we consider a few variables that could have been included in our data set, but were NOT.  Thus, you won't be able to write and test code, but you nevertheless should be able to use your knowledge of the tidyverse to answer these questions.

Here are the hypothetical variables:

- HR_party = party of that county's US Representative (Republican, Democrat, Independent, Green, or Libertarian)
- people_per_MD = number of residents per doctor (higher values = fewer doctors)
- perc_over_65 = percent of residents over 65 years old
- perc_white = percent of residents who identify as white


8. Hypothetical R chunk #1:

```{r}
#| eval: false

# Hypothetical R chunk 1
temp <- vaccine_data |>
  mutate(new_perc_vac = ifelse(perc_complete_vac > 95, NA, perc_complete_vac),
         MD_group = cut_number(people_per_MD, 3)) |>
  group_by(MD_group) |>
  summarise(n = n(),
            mean_perc_vac = mean(new_perc_vac, na.rm = TRUE),
            mean_white = mean(perc_white, na.rm = TRUE))
```  

a) Describe the tibble `temp` created above.  What would be the dimensions?  What do rows and columns represent?

b) What would happen if we replaced `new_perc_vac = ifelse(perc_complete_vac > 95, NA, perc_complete_vac)` with `new_perc_vac = ifelse(perc_complete_vac > 95, perc_complete_vac, NA)`?

c) What would happen if we replaced `mean_white = mean(perc_white, na.rm = TRUE)` with `mean_white = mean(perc_white)`?

d) What would happen if we removed `group_by(MD_group)`?


9. Hypothetical R chunk #2:

```{r}
#| eval: false

# Hypothetical R chunk 2
ggplot(data = vaccine_data) +
  geom_point(mapping = aes(x = perc_over_65, y = perc_complete_vac, 
                           color = HR_party)) +
  geom_smooth()

temp <- vaccine_data |>
  group_by(HR_party) |>
  summarise(var1 = n()) |>
  arrange(desc(var1)) |>
  slice_head(n = 3)

vaccine_data |>
  ggplot(mapping = aes(x = fct_reorder(HR_party, perc_over_65, .fun = median), 
                       y = perc_over_65)) +
    geom_boxplot()
```

a) Why would the first plot produce an error?

b) Describe the tibble `temp` created above.  What would be the dimensions?  What do rows and columns represent?

c) What would happen if we replaced `fct_reorder(HR_party, perc_over_65, .fun = median)` with `HR_party`?


10. Hypothetical R chunk #3:

```{r}
#| eval: false

# Hypothetical R chunk 3
vaccine_data |>
  filter(!is.na(people_per_MD)) |>
  mutate(state_lump = fct_lump(state, n = 4)) |>
  group_by(state_lump, rural_urban_code) |>
  summarise(mean_people_per_MD = mean(people_per_MD)) |>
  ggplot(mapping = aes(x = rural_urban_code, y = mean_people_per_MD, 
      colour = fct_reorder2(state_lump, rural_urban_code, mean_people_per_MD))) +
    geom_line()
```

a) Describe the tibble piped into the ggplot above.  What would be the dimensions?  What do rows and columns represent?

b) Carefully describe the plot created above.

c) What would happen if we removed `filter(!is.na(people_per_MD))`?

d) What would happen if we replaced `fct_reorder2(state_lump, rural_urban_code, mean_people_per_MD)` with `state_lump`?
