---
title: "Strings: Extra Practice (Part 3)"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
---
  
You can download this .qmd file from [here](https://github.com/joeroith/264_spring_2026/blob/main/04_strings_part3.qmd).  Just hit the Download Raw File button.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(rvest)
library(httr)
```


## On Your Own - Extra practice with strings and regular expressions

1. Describe the equivalents of ?, +, * in {m,n} form.

?: {0,1}
+: {1, }
*: {0, }

2. Describe, in words, what the expression `"(.)(.)\\2\\1"` will match, and provide a word or expression as an example.

This will be string that have some 2 charterers and then the second repeats then the first. A example would be "abba" or "noon". 

3. Produce an R string which the regular expression represented by `"\\..\\..\\.."` matches.  In other words, find a string `y` below that produces a TRUE in `str_detect`.

.A.B.C

4. Solve with `str_subset()`, using the words from `stringr::words`:

- Find all words that start or end with x.
```{r}
str_subset(stringr::words, "^x|x$")
```

- Find all words that start with a vowel and end with a consonant.
```{r}
str_subset(stringr::words, "^[aeiou].*[^aeiou]$")
```

- Find all words that start and end with the same letter
```{r}
str_subset(stringr::words, "^(.).*\\1$")
```

5. What words in `stringr::words` have the highest number of vowels? What words have the highest proportion of vowels? (Hint: what is the denominator?)  Figure this out using the tidyverse and piping, starting with `as_tibble(words) |>`.

```{r}
vowel_tabel <- as_tibble(stringr::words) |>
  mutate(vowel_count = str_count(value, "[aeiou]"),
         prop = vowel_count/str_length(value))

vowel_tabel |>
  arrange(-vowel_count)

vowel_tabel |>
  arrange(-prop)
```

6. From the Harvard sentences data, use `str_extract` to produce a tibble with 3 columns:  the sentence, the first word in the sentence, and the first word ending in "ed" (NA if there isn't one).

```{r}
as_tibble(sentences) |>
  mutate(first_word = str_extract(value, "[^ ]+"),
         first_ed = str_extract(value, "\\b[a-zA-Z]*ed\\b"))
```

7. Find and output all contractions (words with apostrophes) in the Harvard sentences, assuming no sentence has multiple contractions.

```{r}
str_extract(sentences, "[a-zA-Z]*\\'[a-zA-Z]*")
```

8. *Carefully* explain what the code below does, both line by line and in general terms.

```{r}
temp <- str_replace_all(words, "^([A-Za-z])(.*)([a-z])$", "\\3\\2\\1")
as_tibble(words) |>
  semi_join(as_tibble(temp)) |>
  print(n = Inf)

```

First we take each word and flip the first and last letter of each word. Then we semi join with the whole words table, which only keeps the words that having matching entires in temp, which is only words that have that word with the first and last letters fliped. This will be words that start and end with the same letter but also words like no and on.