---
title: "Strings: In-class Exercises (Part 2)"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

This uses parts of R4DS Ch 14: Strings and Ch 15: Regular Expressions
(both the first and second editions).

## Manipulating strings

str functions to know for manipulating strings:

-   str_length(): This function gets the length of a string.
-   str_sub(x, y): This gets a slice of the string from x to y.
-   str_c(): This fucntion get
-   str_to_lower() This fucntion turns the string to lowercase.
-   str_to_upper() This funtiion turns the sting to uppercase.
-   str_to_title() This function turns the sting to title case.
-   str_replace() *not in 02_strings examples*

```{r}
#| message: false
#| warning: false
#| results: hide

library(tidyverse)

#spotify <- read_csv("Data/spotify.csv") 
spotify <- 
  read_csv("https://joeroith.github.io/264_spring_2026/Data/spotify.csv")

spot_smaller <- spotify |>
  select(
    title, 
    artist, 
    album_release_date, 
    album_name, 
    subgenre, 
    playlist_name
  )

spot_smaller <- spot_smaller[c(5, 32, 49, 52, 83, 175, 219, 231, 246, 265), ]
spot_smaller
```

## Warm-up

0.  Describe what EACH of the str\_ functions below does. Then, create a
    new variable "month" which is the two digit month from
    album_release_date
    
```{r}
#| eval: false

spot_new <- spot_smaller |>
  select(title, album_release_date) |>
  mutate(title_length = str_length(title), 
         #Get us a new colum of the length of the title.
         year = str_sub(album_release_date, 1, 4),
         # Get a subset of the string release date from 1 to 4 or the year.
         month = parse_number(str_sub(album_release_date, 6,7)), 
         # Get a subset of the string release date from 6 to 7 or the month.
         title_lower = str_to_lower(title),
         # Sets the string to lowercase.
         album_release_date2 = str_replace_all(album_release_date, "-", "/"))
         # Rplaces all of the -'s in the string with /'s
spot_new
         
max_length <- max(spot_new$title_length)

str_c("The longest title is", max_length, "characters long.", sep = " ") 
str_glue("The longest title is {max_length} characters long.")
```

## Important functions for identifying strings which match

str_view() : most useful for testing str_subset() : useful for printing
matches to the console str_detect() : useful when working within a
tibble

1.  Identify the input type and output type for each of these examples:

```{r}
#| eval: false

str_view(spot_smaller$subgenre, "pop")
typeof(str_view(spot_smaller$subgenre, "pop"))
class(str_view(spot_smaller$subgenre, "pop"))

str_view(spot_smaller$subgenre, "pop", match = NA)

str_subset(spot_smaller$subgenre, "pop")

str_detect(spot_smaller$subgenre, "pop")
```

- str_view: The input type is a vector of strings and the output a str view class (weird).
- str_subset: The input is a vector of strings and outputs a vector of strings.
- str_detect: The input is a vector of strings and outputs a logic vector.

2.  Use str_detect to print the rows of the spot_smaller tibble
    containing songs that have "pop" in the subgenre. (i.e. make a new
    tibble with fewer rows)
    
```{r}
(pop_songs <- spot_smaller |>
  filter(str_detect(subgenre, "pop")))
```

3.  Find the mean song title length for songs with "pop" in the subgenre
    and songs without "pop" in the subgenre.
    
```{r}
spot_smaller |>
  mutate(sub_pop = ifelse(str_detect(subgenre,"pop"), 
                          "Genre with pop", "Genre without pop")) |>
  group_by(sub_pop) |>
  summarise(mean_title_length = mean(str_length(title)))
```


4.  In the bigspotify dataset, find the proportion of songs which
    contain "love" in the title (track_name) by playlist_genre.

```{r}
#| message: false
#| warning: false
#| results: hide

bigspotify <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```

```{r}
bigspotify |>
  filter(!is.na(track_name)) |>
  mutate(track_name = str_to_lower(track_name),
         love = str_detect(track_name, "love")) |>
    group_by(playlist_genre) |>
    summarise(prop_love = mean(love))
```


## Matching patterns with regular expressions

|          |                                          |
|----------|------------------------------------------|
| `^abc`   | string starts with exactly "abc"         |
| `abc$`   | string ends with exactly "abc"           |
| `.`      | any character (helpful as a placeholder) |
| `[^abc]` | anything EXCEPT a or b or c              |

```{r}
#| eval: false

# Guess the output!

# Anything that starts with a K.
str_view(spot_smaller$artist, "^K") 

# Anything that was relased on the first day of the month. 
str_view(spot_smaller$album_release_date, "01$")

# Any title that has 2 charters then a space.
str_view(spot_smaller$title, "^.. ")

# Any artist that does not have capital, lowercase, or spaces
str_view(spot_smaller$artist, "[^A-Za-z ]")
```

5.  Given the corpus of common words in stringr::words, create regular
    expressions that find all words that:

-   Start with “y”.
-   End with “x”
-   Are exactly three letters long.
-   Have seven letters or more.
-   Start with a vowel.
-   End with ed, but not with eed.
-   Words where q is not followed by u. (are there any in `words`?)

```{r}
# Try using str_view() or str_subset()
str_subset(stringr::words, "^y")
str_subset(stringr::words, "x$")
str_subset(stringr::words, "^...$")
str_subset(stringr::words, "^.{7,}$")
str_subset(stringr::words, "^[aeiou]")
str_subset(stringr::words, "[^e]ed$")
str_subset(stringr::words, "q[^u]")
```

## More useful regular expressions:

|       |                              |
|-------|------------------------------|
| `\\d` | any number                   |
| `\\s` | any space, tab, etc.         |
| `\\b` | any boundary: space, ., etc. |

```{r}
str_view(spot_smaller$album_name, "\\d")
str_view(spot_smaller$album_name, "\\s")
str_view(spot_smaller$album_name, "\\b")
```

Here are the regular expression special characters that require an
escape character (a preceding `\`): `^` `$` `.` `?` `*` `|` `+` `(` `)`
`[` `{`

For any characters with special properties, use `\` to "escape" its
special meaning ... but `\` is itself a special character ... so we need
two `\\`! (e.g. `\\$`, `\\.`, etc.)

```{r}
str_view(spot_smaller$title, "$")
str_view(spot_smaller$title, "\\$")
```

6.  In bigspotify, how many track_names include a `$`? Be sure you print
    the track_names you find and make sure the dollar sign is not just
    in a featured artist!

```{r}
bigspotify |>
filter(str_detect(track_name, "\\$")) |>
  filter(!str_detect(track_name, "(feat|with).*\\$")) |>
  select(track_name, track_artist)
  
```
    

7.  In bigspotify, how many track_names include a dollar amount (a `$`
    followed by a number).
    
```{r}
bigspotify |>
  filter(str_detect(track_name, "\\$\\d+")) |>
  select(track_name, track_artist)
```
    

## Repetition

|         |                       |
|---------|-----------------------|
| `?`     | 0 or 1 times          |
| `+`     | 1 or more             |
| `*`     | 0 or more             |
| `{n}`   | exactly n times       |
| `{n,}`  | n or more times       |
| `{,m}`  | at most m times       |
| `{n,m}` | between n and m times |

```{r}
str_view(spot_smaller$album_name, "[A-Z]{2,}")

str_view(spot_smaller$album_release_date, "\\d{4}-\\d{2}")
```

**Use at least 1 repetition symbol when solving 8-10 below**

8.  Modify the first regular expression above to also pick up "A.A" (in
    addition to "BEYONC" and "II"). That is, pick up strings where there
    might be a period between capital letters.
    
```{r}
str_view(spot_smaller$album_name, "[A-Z.]{2,}")
```
  

9.  Create some strings that satisfy these regular expressions and
    explain.

-   `"^.*$"`: This string is any string that from the start to the end it start with any character then repeats it 0 or more times. so this is basically any character.
```{r}
str_view(c("a","Henry","Test",""), "^.*$")
```

-   `"\\{.+\\}"`: This is strings that have \{\} and any character repeated 1 or more times.
```{r}
str_view(c("{aa}","aa","{a}","{}"), "\\{.+\\}")
```


10. Create regular expressions to find all `stringr::words` that:

-   Start with three consonants.
```{r}
str_view(stringr::words, "^[^aeiou]{3}")
```
-   Have two or more vowel-consonant pairs in a row.
```{r}
str_view(stringr::words, "([aeiou][^aeiou]){2,}")
```

## Useful functions for handling patterns

str_extract() : extract a string that matches a pattern str_count() :
count how many times a pattern occurs within a string

```{r}
str_extract(spot_smaller$album_release_date, "\\d{4}-\\d{2}")

spot_smaller |>
  select(album_release_date) |>
  mutate(year_month = str_extract(album_release_date, "\\d{4}-\\d{2}"))

spot_smaller |>
  select(artist) |>
  mutate(n_vowels = str_count(artist, "[aeiou]"))
```

11. In the spot_smaller dataset, how many words are in each title?
    (hint: `\\b`)
    
```{r}
spot_smaller |>
  mutate("Number_of_words" = str_count(title, "\\b")/2) |>
  select(title, Number_of_words)
```
    
12. In the spot_smaller dataset, extract the first word from every
    title. Show how you would print out these words as a vector and how
    you would create a new column on the spot_smaller tibble.
    
```{r}
str_extract(spot_smaller$title, "^[^ ]+")
```

```{r}
spot_smaller |>
  mutate(first_word = str_extract(title, "^[^ ]+")) |>
  select(title, first_word)
```

13. Which decades are popular for playlist_names? Using the bigspotify
    dataset, try doing each of these steps one at a time!

-   filter the bigspotify dataset to only include playlists that include
    something like "80's" or "00's" in their title.
-   create a new column that extracts the decade
-   use count to find how many playlists include each decade
-   what if you include both "80's" and "80s"?
-   how can you count "80's" and "80s" together in your final tibble?

```{r}
bigspotify |>
  filter(str_detect(playlist_name, "\\d{2}'?s")) |>
  filter(!str_detect(playlist_name, "808's")) |> # This is to remove the 808's
  mutate(decade = str_extract(playlist_name, "\\d{2}'?s"),
        decade = str_replace(decade, "'", "")) |>
  count(decade, sort = TRUE)
```


## Grouping and backreferences

```{r}
# find all fruits with repeated pair of letters.  
fruit = stringr::fruit
fruit
str_view(fruit, "(..)\\1", match = TRUE)

# why does the code below add "pepper" and even "nectarine"?
str_view(fruit, "(..)(.*)\\1", match = TRUE)
```

Tips with backreference:

-   You must use () around the the thing you want to reference.
-   To backreference multiple times, use `\\1` again.
-   The number refers to which spot you are referencing... e.g. `\\2`
    references the second set of ()

```{r}
x1 <- c("abxyba", "abccba", "xyaayx", "abxyab", "abcabc")
str_view(x1, "(.)(.)(..)\\2\\1")
str_view(x1, "(.)(.)(..)\\1\\2")
str_view(x1, "(.)(.)(.)\\1\\2\\3")
```

14. Describe to your groupmates what these expressions will match, and
    provide a word or expression as an example:

-   `"(.)\\1\\1"` : This will be 3 repeated charatures. "aaa"  
-   `"(.)(.)(.).*\\3\\2\\1"`: Any string that starts with three characters and later contains those same three characters in reverse order. "abcdefgcba".

Which words in `stringr::words` match each expression?

15. Construct a regular expression to match words in `stringr::words`
    that contain a repeated pair of letters (e.g. “church” contains “ch”
    repeated twice) but *not* match repeated pairs of numbers (e.g.
    507-786-3861).
    
```{r}
str_view(stringr::words, "([a-z]{2})\\1")
```
    
16. Reformat the album_release_date variable in spot_smaller so that it
    is MM-DD-YYYY instead of YYYY-MM-DD. (Hint: str_replace().)
    
```{r}
spot_smaller |>
  mutate(album_release_date = str_replace(
    album_release_date,
    "(\\d{4})-(\\d{2})-(\\d{2})",
    "\\2-\\3-\\1"
  ))
```


17. BEFORE RUNNING IT, explain to your partner(s) what the following R
    chunk will do:

```{r}
sentences |> 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") |> 
  head(5)
```

This takes the second and 3rd word and flips them.
