---
title: "Patchwork"
subtitle: "Henry Norton-Bower"
format:
  revealjs: 
    theme: solarized
    chalkboard: 
      buttons: false
    preview-links: auto
---

```{r}
#| include: false
library(tidyverse)
library(ggthemes)
library(patchwork)
```

## What is Patchwork?

::: incremental
- A package to combine ggplot's into one nice looking figure.
- Built on top of ggplot and works with any ggplot object
- Uses intuitive operators like `+`, `/`, and `|` to arrange plots
- Handles alignment, spacing, and shared legends automatically
- Great for making papers look better and more clean
:::

# How do each of the operators work?

---

## The `+` Operator

- places ggplot objects next to each other in a grid.

```{r} 
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
#| output: false
#| code-line-numbers: "0|12"

fig1 <- penguins |>
  ggplot(aes(x = species, y = body_mass, fill = species)) +
  geom_boxplot() +
  theme_bw()

fig2 <- penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x=sex, fill = sex)) +
  geom_bar() +
  theme_bw()

fig1+fig2
```

---

## Example of the `+` Operator

```{r}
#| echo: false
fig1+fig2 
```

---

## Example of the `+` Operator

And Patchwork will auto wrap the row for us.

```{r}
#| echo: false
fig3 <- penguins |>
  ggplot(aes(x=island, fill = island)) +
  geom_bar() +
  theme_bw()


fig4 <- penguins |>
  ggplot(aes(x = species, y = flipper_len, fill = species)) +
  geom_boxplot() +
  theme_bw()
```

```{r}
#| echo: true
#| output: false
#| code-line-numbers: "1"
fig1+fig2+fig3+fig4
```

::: {.fragment .fade-in}
```{r}
fig1+fig2+fig3+fig4
```
:::

---

## The `/` operator

- stacks ggplot objects on top of each other rather then in a grid and can be combined with the `+`


```{r} 
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
#| output: false
#| code-line-numbers: "0|7"

(fig1+fig2)/fig3
```


---

## Example of the `/` Operator

```{r}
(fig1+fig2)/fig3
```

---

## The `|` operator
- places ggplot objects explicitly side-by-side, preventing auto-wrapping which is useful when combined with `/` for precise layout control.
```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
#| output: false
fig1 | (fig2 / fig3)
```
---

## Example of the `|` Operator

```{r}
fig1 | (fig2 / fig3)
```


# Extra Styling Options

---

## `plot_layout()`

- Defines the structure of the layout grid: number of rows, columns, how plots fill it, and shared legends.

::: incremental
- `ncol`: number of objects per column.
- `nrow`: number of objects per rows. 
- `byrow`: True by default, if false the plots will be filled in in column order.
- `widths`/`heights`: The widths and heights of each row and columns.  
- `guides`: set to "collect" if you want shared legend
:::


---

## `plot_layout()`

::: {.r-stack}

::: {.fragment .fade-in-then-out}
```{r}
#| echo: true
fig1+fig2+fig3
```
:::

::: {.fragment .fade-in-then-out}
```{r}
#| echo: true
fig1+fig2+fig3 + plot_layout(nrow = 2)
```
:::

::: {.fragment .fade-in-then-out}
```{r}
#| echo: true
fig1+fig2+fig3+fig4+ plot_layout(nrow = 2, 
                             byrow = FALSE, 
                             guides = "collect") 
```
:::

:::

---

## `plot_annotation()`

::: {.r-stack}

::: {.fragment .fade-in-then-out style="position: absolute; top: 100px; left: 0;"}
- Can be used to add a title, subtitle, or caption to the figure.
:::

::: {.fragment .fade-in-then-out}
```{r}
#| echo: true
fig1 + fig2 + plot_layout(guides = "collect") +
  plot_annotation(title = "Combined Penguin Figures",
                  caption = "Data: palmerpenguins R package",
                  theme = theme(
                    plot.title = element_text(size = 16, face = "bold")))
```
:::

:::

---

## `guide_area()`:

This can be added to set aside area for the guide of the figure.

```{r}
#| echo: true
guide_area() + fig1 + fig2 + 
  plot_layout(guides = "collect") 
```


---

## Something to note 

- Even after combining plots with patchwork, you can still modify individual ggplot objects using `+` as usual and patchwork won't interfere.
```{r}
#| echo: true
fig1 + theme_classic() + fig2 + theme_dark()
```

---
## One Last Graph

```{r}
p1 <- penguins |>
  drop_na() |> # get rid of all NA
  ggplot(aes(x = flipper_len, y = body_mass, color = species)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- penguins |>
  drop_na() |>
  ggplot(aes(x = species, y = bill_len, fill = species)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) + # Makes a little cute box plot in the violin plot
  labs(x = NULL, y = "Bill length (mm)") +
  theme_minimal() 

p3 <- penguins |>
  drop_na() |>
  ggplot(aes(x = bill_len, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(x = "Bill length (mm)", y = "Density", fill = "Species") +
  theme_minimal() +
  theme(legend.position = "none")

p4 <- penguins |>
  drop_na() |>
  ggplot(aes(x = island, fill = species)) +
  geom_bar(position = "fill") +
  labs(x = NULL, y = "Proportion", fill = "Species") +
  scale_y_continuous(labels = scales::percent) + # makes the scale percent :D
  theme_minimal() +
  theme(legend.position = "none")

(p1 | p2) / (p4 | p3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Penguins: Body Characteristics Across Species",
    caption = "Data: palmerpenguins R package",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold"))
  )
```



